---
title: 'SHARP Multiomics Workshop Session 2: Interactions'
author: "Andre Kim"
date: "`r Sys.time()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: show
  pdf_document: default
---

<!-- ```{css, echo=FALSE} -->
<!-- pre { -->
<!--   max-height: 200px; -->
<!--   overflow-y: auto; -->
<!-- } -->
<!-- ``` -->


```{r setup, include=FALSE, echo=FALSE}
library(tidyverse)
library(MultiAssayExperiment)
library(knitr)
library(ggplot2)
library(reshape2)
library(gap)
library(kableExtra)

options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE)

# suppress warning messages for final rendering
old.warn <- getOption("warn")
options(warn=-1)
```


# SHARP MULTIOMICS WORKSHOP

## Data Descriptions for Interactions

The HELIX study represents a collaborative project across six established and ongoing longitudinal population-based birth cohort studies in six European countries (France, Greece, Lithuania, Norway, Spain, and the United Kingdom). HELIX used a multilevel study design with the entire study population totaling 31,472 mother–child pairs, recruited during pregnancy, in the six existing cohorts (first level); a subcohort of 1301 mother-child pairs where biomarkers, omics signatures and child health outcomes were measured at age 6-11 years (second level); and repeat-sampling panel studies with around 150 children and 150 pregnant women aimed at collecting personal exposure data (third level). For more details on the study design see Vrijheid, Slama, et al. EHP 2014. see https://www.projecthelix.eu/index.php/es/data-inventory for more information regarding the study.

In this example, we analyze data from 1,122 children 6-11 years old. 

### Variables:  
* **hs_bmi_c_cat (Outcome)** - Body mass index
  + 0 = Thinness/Normal
  + 1 = Overweight/Obese (WHO reference)
* **lifestyles_e3_alcpreg_yn_None (Interaction exposure)** - Maternal alcohol use during pregnancy
  + 0 = No
  + 1 = Yes
* **h_mbmi_None** - Maternal pre-pregnancy body mass index
  + kg/m2 (continuous)
* **covariates_e3_sex_None** - Sex
  + 0 = Female
  + 1 = Male
* **h_age_None** - Maternal age
  + years (continuous)
* **h_cohort** - Country
  + 1 = France 
  + 2 = Greece
  + 3 = Lithuania
  + 4 = Norway
  + 5 = Spain 
  + 6 = UK
* **h_edumc_None** - Maternal education
  + 1 = primary school
  + 2 = secondary school
  + 3 = university degree or higher
* **ethn_PC1/2** - principal components (ethnicity)
  

dataset <- data.frame(d[,c("primary", "hs_bmi_c_cat", 
"covariates_h_mbmi_None", "covariates_e3_sex_None", "covariates_h_age_None", 
"covariates_h_cohort", "covariates_h_edumc_None", "proteome.cov_ethn_PC1", 
"proteome.cov_ethn_PC2", "proteome.cov_h_ethnicity_cauc", "metabol_urine.cov_ethn_PC1", 
"metabol_urine.cov_ethn_PC2", "metabol_urine.cov_h_ethnicity_cauc", 
"metabol_serum.cov_ethn_PC1", "metabol_serum.cov_ethn_PC2", "metabol_serum.cov_h_ethnicity_cauc")])



<br>

```{r GxE: Data Analysis setup, echo=T }
# Outcome
outcome.Name <- "hs_bmi_c_cat"

# Exposure variables
# exposure.Name <- "e3_asmokcigd_p_None"
exposure.Name <- "e3_alcpreg_yn_None"

covariate.Names <- c("h_mbmi_None","e3_sex_None","h_age_None","h_cohort","h_edumc_None","ethn_PC1","ethn_PC2")

# SNPs
snp.Names <- paste("SNP", 1:1000, sep=".")

# Analysis models to run
GxE.univariate <- T
```


## GxE overview

### The question of interest:
- Is there effect heterogeneity for the association of **`r exposure.Name`** with the outcome **`r outcome.Name`** by germline genetics?  
 
<br>

### Interaction model

## GxE: Processing the Data

`r MultiAssayExperiment` is an R object for storing multiple datasets measured on the same individual. We'll start by specifying and



```{r GxE: Processing the Data, echo=TRUE, cache = T}
load("~/Google Drive/My Drive/SHARP.MultiomicsWorkshop.USC.share/data/HELIX.MultiAssayExperiment.RData")
variables <- c(covariate.Names, exposure.Name, "h_ethnicity_cauc", snp.Names)






# 1) select variables but keep in MultiAssayExperiment format
# 2) intersectionColumns selects only individuals with complete data
# 3) wideFormat returns as a DataFrame
# d <- wideFormat(intersectColumns(helix_ma[variables, ,]), colDataCols=outcome.Name) # comment out for now
d <- readRDS("data/d_alcohol.rds")

# Create design matrix
X <- d[,grep("SNP", names(d))]
names(X) <- snp.Names
X <- as.matrix(X)

# Create the outcome variable
Y <- d[,outcome.Name] # outcome
if(outcome.Name=="hs_bmi_c_cat") { Y <- ifelse(as.numeric(Y)>=3, 1, 0)}

# Create the covariate design matrix
U <- d[,c(paste0("covariates_", covariate.Names[1:5]), paste0("proteome.cov_", covariate.Names[6:7]))]
names(U) <- covariate.Names
U[,c("h_cohort","e3_sex_None","h_edumc_None")] <- lapply(U[,c("h_cohort","e3_sex_None","h_edumc_None")], factor)
U[,c("h_mbmi_None", "h_age_None","ethn_PC1","ethn_PC2")] <- lapply(U[,c("h_mbmi_None", "h_age_None","ethn_PC1","ethn_PC2")], as.numeric)
U <- model.matrix(as.formula(paste("~-1+", paste(covariate.Names, collapse="+"))), data=U)
U = U[,-which(colnames(U) == "e3_sex_Nonemale")]

# Other variables for analysis
N <- nrow(d) # number of individuals in the analysis
Q <- ncol(U)  # number of covariates in the matrix U
P <- ncol(X)  # number of SNPs in the matrix X

# Exposure variable
E <- as.numeric(d[,grep(exposure.Name, names(d))])  
E <- ifelse(E > 0,1,0) # Dichotomize
```
<br>

## GxE: Descriptive Statistics



### Exposure data

Descriptive statistics 

```{r echo = T, message=F}
library(table1)

# temporary dataset for descriptive table and univariate analysis
dataset <- data.frame(d[,c("primary", "hs_bmi_c_cat", "lifestyles_e3_alcpreg_yn_None", 
"covariates_h_mbmi_None", "covariates_e3_sex_None", "covariates_h_age_None", 
"covariates_h_cohort", "covariates_h_edumc_None", "proteome.cov_ethn_PC1", 
"proteome.cov_ethn_PC2", "proteome.cov_h_ethnicity_cauc", "metabol_urine.cov_ethn_PC1", 
"metabol_urine.cov_ethn_PC2", "metabol_urine.cov_h_ethnicity_cauc", 
"metabol_serum.cov_ethn_PC1", "metabol_serum.cov_ethn_PC2", "metabol_serum.cov_h_ethnicity_cauc")])
dataset$covariates_h_mbmi_None <- as.numeric(dataset$covariates_h_mbmi_None)
dataset$lifestyles_e3_alcpreg_yn_None <- factor(dataset$lifestyles_e3_alcpreg_yn_None, labels = c("No", "Yes"))
dataset$covariates_e3_sex_None <- factor(dataset$covariates_e3_sex_None, labels = c("Female", "Male"))
dataset$hs_bmi_c_cat <- factor(ifelse(as.numeric(dataset$hs_bmi_c_cat)>=3, 1, 0), label = c("Thinness/Normal", "Overweight/Obese"))
dataset$covariates_h_edumc_None <- factor(dataset$covariates_h_edumc_None, labels = c("Primary school", "Secondary school", "University degree or higher"))

label(dataset$lifestyles_e3_alcpreg_yn_None) <- "Maternal alcohol during pregnancy"
label(dataset$covariates_e3_sex_None) <- "Child sex"
label(dataset$covariates_h_mbmi_None) <- "Maternal pre-pregnancy BMI"
label(dataset$covariates_h_edumc_None) <- "Maternal education"


table1(~ lifestyles_e3_alcpreg_yn_None + 
         covariates_e3_sex_None + 
         covariates_h_edumc_None +
         covariates_h_mbmi_None | hs_bmi_c_cat, data=dataset)

```

<br>
<br>

( need to rework univariate analysis , will show E + covariates)

```{r, echo = T, results='asis', message = F}
# note: uses tidyverse package

# run variable | outcome regression models (univariate)

univariate_vars <- c("lifestyles_e3_alcpreg_yn_None", "covariates_e3_sex_None", "covariates_h_edumc_None", "covariates_h_mbmi_None")









# univariate_vars <- c("lifestyles_e3_alcpreg_yn_None")
#
# p <- univariate_vars

univariate.results <- t(sapply(univariate_vars, FUN=function(p) {  # using index p facilitate write
  x <- dataset[,p]
  reg <- glm(dataset$hs_bmi_c_cat ~ x, family=binomial)    # perform logistic regression
  s.reg <- summary(reg)                 # get the summary for the regression
  c.reg <- s.reg$coef[2,]             # select the coefficients for the exposure
  return(c.reg)                         # to avoid potential memory issues only return coefficients if small number of exposures
}, simplify=T))
univariate.results <- data.frame(univariate_vars,univariate.results)
names(univariate.results) <- c("Protein","Estimate", "SD","Z.Statistic", "P-value")
univariate.results$`P-value` <-  format(univariate.results$`P-value`, scientific=T)

```



### Genomic data {.tabset}

- The genome includes a total of `r P` single nucleotide polymorphisms (SNPs):

#### Plot of Genetic Ancestry as Estimated by Prinicpal Components (included in PreWorkshop Lab)
```{r Genome: PC plot, echo=TRUE}
plot(d$proteome.cov_ethn_PC1, d$proteome.cov_ethn_PC2, pch=16, col=ifelse(d$proteome.cov_h_ethnicity_cauc=="yes", 1, 2),
     xlab="Component 1", ylab="Component 2")
legend(x="topleft", legend=c("Caucasian", "Other"), col=c(1,2), pch=16)
```

#### Correlation Matrix for Local Region of the Genome:
```{r Genome: cor.plot, echo=TRUE}
cormat <- round(cor(X[,1:(P/5)], use="complete.obs"), 2)
cormat[lower.tri(cormat)]<- NA
melted_cormat <- melt(cormat)
ggplot(data = melted_cormat, aes(Var2, Var1, fill = value))+
  geom_tile(color = "white")+
  scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                       midpoint = 0, limit = c(-1,1), space = "Lab",
                       name="Pearson\nCorrelation") +
  theme_minimal()+
  theme(axis.text.x = element_blank(), axis.text.y = element_blank())+
  labs(y= "SNPs", x = "SNPs")+
  coord_fixed()

```
<br>

## GxE: Statistical Approaches for Association Analysis
### GxE Univariate Regression

```{r GxE: Univariate model, echo=TRUE}
if(GxE.univariate) {
  int.results <- t(sapply(1:ncol(X), FUN=function(p) {  # using index p faciliate write
    g <- X[,p]
    reg <- glm(Y~g*E+U, family=binomial)    # perform logistic regression
    s.reg <- summary(reg)                 # get the summary for the regression
    c.reg <- as.numeric(s.reg$coef["g:E",])             # select the coefficients
    # write.table(t(c(paste("SNP", p, sep="."), c.reg)), file="IntResultsGenome.txt", append=ifelse(p==1, F, T), quote=F, sep="\t", col.names=ifelse(p==1, T, F), row.names=F)
    return(c.reg)                         # to avoid potential memory issues only return coefficients if small number of exposures
  }, simplify=T))
  int.results <- data.frame(paste("E:SNP", 1:ncol(X), sep="."),int.results)
  names(int.results) <- c("E:SNP.Name","Estimate", "SD","Z.statistic", "P.value")
}
```

#### GxE Univariate results: {.tabset}
##### GxE Univariate Summary Table:
```{r GxE: Univariate table}
if(GxE.univariate) { kable(int.results[int.results$P.value<0.05,], digits=3, align="c", row.names=FALSE, col.names=c("E:SNP","Estimate", "SD","Z statistic", "P value"))}
```

##### GxE Univariate Manhattan Plot:
```{r GxE: Univariate MH plot}
neglog.pvalues <- -log10(int.results$P.value)
plot(1:nrow(int.results), neglog.pvalues,
     pch=16, xaxt="n", ylim=c(0, max(neglog.pvalues, 3)),
     ylab="-log(p-value)", xlab="SNPs")
abline(h=-log10(0.05/nrow(int.results)), lty=2, lwd=2, col=2)
```

##### GxE Univariate QQ-Plot:
```{r GxE: QQ-plot}
pvalues <- int.results$P.value
r <- gcontrol2(pvalues, pch=16)
lambda <- round(r$lambda,3)
text(x=1, y=5, labels=bquote(lambda == .(lambda)), cex=2)
```

<br>

```{r final clean up}
options(warn=old.warn)

```





# GxEScanR

https://cran.r-project.org/web/packages/GxEScanR/index.html
https://cran.r-project.org/web/packages/BinaryDosage/index.html


## Introduction


### BinaryDosage
Genotype imputation is an essential tool in genomics, enabling association testing with markers not directly genotyped, increasing statistical power, and facilitating data pooling between studies that employ different genotyping platforms. Two commonly used software packages for imputation are minimac and Impute2. Furthermore, services such as the Michigan Imputation Server have made genotype imputation much more accessible and streamlined.

While a number of software options are available for analyses of imputed data (e.g. PLINK, EPACTS), fewer are available for Genomewide Gene x Environment Interaction Scan (GWIS). Furthermore, data management tasks such as parsing, subsetting, and merging, while manageable in smaller studies, quickly become unwieldy and prohibitively slow with very large samples sizes. We aim to address these limitations by converting imputation outputs into a binary dosage file. The benefits of a binary format are two fold - decreased hard drive storage requirements (compared to a VCF file), and speed of parsing/analyses. The **BinaryDosage** package contains functions to convert VCF and Impute2 formatted files into binary dosage files, along with functions to merge samples.


### GxEScanR
**GxEScanR** is designed to efficiently run genome-wide association study (GWAS) and genome-wide by environmental interaction study (GWEIS) scans using imputed genotypes stored in the BinaryDosage format. The phenotype to be analyzed can either be a continuous or binary trait. The GWEIS scan performs multiple tests that can be used in two-step methods.


## Analysis


### Data preparation

BinaryDosage requires imputed genotype files as input. Generate VCF and INFO files (e.g. Minimac output)

```{r, echo = T, eval = F}
library(BinaryDosage)

# Create VCF file
vcf <- data.frame(t(X)) # transpose

# enter dummy values for required fields
vcf$chr <- "chr1"
vcf$pos <- seq(1,1000)
vcf$ID <- snp.Names
vcf$FORMAT <- "DS"
vcf[c('REF', 'ALT', 'QUAL', 'FILTER', 'INFO', 'Sample')] <- rep('.', nrow(vcf))

vcf <- vcf[, c('chr', 'pos', 'ID', 'REF', 'ALT', 'QUAL', 'FILTER', 'INFO', 'FORMAT')]
colnames(vcf) <- c('#CHROM', 'POS', 'ID', 'REF', 'ALT', 'QUAL', 'FILTER', 'INFO', 'FORMAT')
vcfout <- cbind(vcf, data.frame(t(X)))
write.table(vcfout, "./data/out.vcf", quote = F, row.names = F, col.names = T, sep = "\t")


# Create INFO file
info <- vcf[, c('ID', 'REF', 'ALT')]
info[c('AAF', 'MAF', 'AvgCall', 'Rsq')] <- rep(0.9, nrow(vcf))
info$Genotyped = "imputed"
info[c('LooRsq', 'EmpR', 'EmpRsq', 'Dose0', 'Dose1')] = "-"
info <- info[, c('ID', 'REF', 'ALT', 'AAF', 'MAF', 'AvgCall', 'Rsq', 'Genotyped', 'LooRsq', 'EmpR', 'EmpRsq', 'Dose0', 'Dose1')]
colnames(info) <- c('SNP', 'REF(0)', 'ALT(1)', 'ALT_Frq', 'MAF', 'AvgCall', 'Rsq', 'Genotyped', 'LooRsq', 'EmpR', 'EmpRsq', 'Dose0', 'Dose1')

write.table(info, "./data/out.info", quote = F, row.names = F, col.names = T, sep = "\t")

# output BinaryDosage file
BinaryDosage::vcftobd(vcffiles = c("./data/out.vcf", './data/out.info'), gz = FALSE, bdfiles = "./data/out.bdose")
```

## Run GxEScanR

Requires external covariate file - ID, Outcome, C, E (interaction variable must be in last position)

```{r, echo = T}
library(BinaryDosage)
library(GxEScanR)

# read BinaryDosage info
bdinfo <- BinaryDosage::getbdinfo("data/out.bdose")

# Create covariate file
covar <- data.frame(U)
covar$outcome = Y
# covar$e3_asmokcigd_p_None = E
covar$e3_alcpreg_yn_None = E
covar$ID = paste0("X", seq(1, 1122))

# covar <- covar[, c("ID", "outcome", "h_mbmi_None", "e3_sex_Nonefemale", "h_age_None", "h_cohort2", "h_cohort3", "h_cohort4", "h_cohort5", "h_cohort6", "h_edumc_None2", "h_edumc_None3", "ethn_PC1", "ethn_PC2", "e3_asmokcigd_p_None")]
covar <- covar[, c("ID", "outcome", "h_mbmi_None", "e3_sex_Nonefemale", "h_age_None", "h_cohort2", "h_cohort3", "h_cohort4", "h_cohort5", "h_cohort6", "h_edumc_None2", "h_edumc_None3", "ethn_PC1", "ethn_PC2", "e3_alcpreg_yn_None")]
write.table(covar, "./data/covar.txt", row.names = F, quote = F, sep = '\t')

# Run GxEScanR
output <- GxEScanR::gweis(data = covar, bdinfo = bdinfo)

# create EDGE statistic (used in two-step method)
output$lrtedge = output$lrtdg + output$lrteg
```


## Q-Q plots {.tabset}
```{r, echo = T}
# qq plot function

plot_qq <- function(data, statistic, df) {
  pvalues <- pchisq(data[, statistic], df = df, lower.tail = F)
  r <- gcontrol2(pvalues, pch=16)
  lambda <- round(r$lambda,3)
  text(x=1, y=5, labels=bquote(lambda == .(lambda)), cex=2)
}
```

### D|G
```{r, echo = T}
plot_qq(output, 'lrtdg', 1)
```

### GxE
```{r, echo = T}
plot_qq(output, 'lrtgxe', 1)
```

### Case only
```{r, echo = T}
plot_qq(output, 'lrtcase', 1)
```

### 2DF
```{r, echo = T}
plot_qq(output, 'lrt2df', 2)
```

### 3DF
```{r, echo = T}
plot_qq(output, 'lrt3df', 3)
```





## Manhattan plots {.tabset}

```{r echo = T}
# manhattan plot

data = output
statistic = 'lrtcase'; df = 1

plot_manhattan <- function(data, statistic, df) {
  
  # remove NA reults if any
  data <- data[which(!is.na(data[,statistic])), ]
  
  # plot
  neglog.pvalues <- -log10(pchisq(data[,statistic], lower.tail = F, df = df))
  plot(1:nrow(data), neglog.pvalues,
       pch=16, xaxt="n", ylim=c(0, max(neglog.pvalues, 3)),
       ylab="-log(p-value)", xlab="SNPs")
  abline(h=-log10(0.05/nrow(data)), lty=2, lwd=2, col=2)
}

# significant findings
sig_table <- function(data, statistic, df) {
  varname = paste0(statistic, "_pval")
  data[varname] <- pchisq(data[,statistic], lower.tail = F, df = df)
  out <- data[which(data[,varname] < (0.05 / nrow(data))), ]
  return(out[, c("snp", varname)])
}

```

### D|G
```{r, echo = T}
plot_manhattan(output, 'lrtdg', 1)
```

### GxE
```{r, echo = T}
plot_manhattan(output, 'lrtgxe', 1)
# kable(sig_table(output, 'lrtgxe', 1))
```

### Case-only
```{r, echo = T}
plot_manhattan(output, 'lrtcase', 1)
```

### 2DF
```{r, echo = T}
plot_manhattan(output, 'lrt2df', 2)
```

### 3DF
```{r, echo = T}
plot_manhattan(output, 'lrt3df', 3)
```

## Two-step methods {.tabset}
```{r echo = T}
# twostep helper function
plot_twostep <- function(data, step1_statistic, step1_df, sizeBin0, alpha) {

  # create bins
  m = nrow(output)
  nbins = ceiling(log2(m/sizeBin0 + 1))
  sizeBin = c(sizeBin0 * 2^(0:(nbins-2)), m - sizeBin0 * (2^(nbins-1) - 1) )
  endpointsBin = cumsum(sizeBin)

  rk.pv <- c(1:m)
  grp <- ceiling(log(rk.pv/sizeBin0+1,base=2))
  rep_helper <- c(table(grp))
  alphaBin = alpha * 2 ^ -(1:nbins) / sizeBin
  alphaBin_dat <- rep(alphaBin, rep_helper)

  # create pvalues and bin variables
  data$step1p <- pchisq(data[, step1_statistic], df = step1_df, lower.tail = F)
  data$step2p <- pchisq(data[, 'lrtgxe'], df = 1, lower.tail = F)
  data                   <- data[order(-data[,step1_statistic]), ]   # sort by step1_statistic
  data$bin_number        <- as.numeric(grp)
  data$step2p_siglvl     <- as.numeric(alphaBin_dat)
  data$log_step2p_siglvl <- -log10(data$step2p_siglvl)
  data$log_step2p        <- -log10(data$step2p)

  significant_hits <- data[which(data$step2p < data$step2p_siglvl), ]

  # create plot
  # split data into lists (for each bin)
  data_plot <- split(data, f = list(data$bin_number))

  # mapinfo just to make sure SNPs are plotted evenly in each bin
  create_mapinfo <- function(x) {
    mapinfo <- seq(unique(x$bin_number) - 1 + 0.1, unique(x$bin_number) - 1 + 0.9, length.out = nrow(x))
    out <- cbind(x, mapinfo)
    return(out)
  }

  data_plot <- lapply(data_plot, create_mapinfo)
  logp_plot_limit = 12

  # PLOT
  binsToPlot = length(data_plot)
  color <- rep(c("#377EB8","#4DAF4A"),100)
  par(mar=c(6, 7, 6, 3))

  bin_to_plot = data_plot[[1]]
  plot(bin_to_plot$mapinfo, bin_to_plot$log_step2p,
       col = ifelse(bin_to_plot$snp %in% significant_hits[, 'snp'], '#E41A1C','#377EB8'),
       pch = ifelse(bin_to_plot$snp %in% significant_hits[, 'snp'], 19, 20),
       cex = ifelse(bin_to_plot$snp %in% significant_hits[, 'snp'], 1.3, 1.7),
       xlab="Bin number for step1 p value",
       ylab="-log10(step2 chiSqGxE p value)",
       xlim=c(0, binsToPlot),
       ylim=c(0, logp_plot_limit),
       axes=F,
       cex.main = 1.7,
       cex.axis = 1.7,
       cex.lab = 1.7,
       cex.sub = 1.7)
  lines(bin_to_plot$mapinfo, bin_to_plot$log_step2p_siglvl, col = "black", lwd=1)

  # remaining bins
  for(i in 2:binsToPlot) {
    bin_to_plot = data_plot[[i]]
    points(bin_to_plot$mapinfo, bin_to_plot$log_step2p,
           col = ifelse(bin_to_plot$snp %in% significant_hits$snp, '#E41A1C', color[i]),
           pch = ifelse(bin_to_plot$snp %in% significant_hits$snp, 19, 20),
           cex = ifelse(bin_to_plot$snp %in% significant_hits$snp, 1.3, 1.7),
           cex.main = 1.7,
           cex.axis = 1.7,
           cex.lab = 1.7,
           cex.sub = 1.7)
    lines(bin_to_plot$mapinfo, bin_to_plot$log_step2p_sig, col = "black",lwd = 1)
  }

  axis(1, at = c(-1.5, seq(0.5, binsToPlot-0.2, 1)), label = c(0, seq(1, binsToPlot, 1)), cex.axis = 1.7)
  axis(2, at = c(0:floor(logp_plot_limit)), label = c(0:logp_plot_limit), cex.axis=1.7)
}


# significant findings
# report SNP, step1, step2
sig_twostep <- function(data, step1_statistic, step1_df, sizeBin0, alpha) {

  # create bins
  m = nrow(output)
  nbins = ceiling(log2(m/sizeBin0 + 1))
  sizeBin = c(sizeBin0 * 2^(0:(nbins-2)), m - sizeBin0 * (2^(nbins-1) - 1) )
  endpointsBin = cumsum(sizeBin)

  rk.pv <- c(1:m)
  grp <- ceiling(log(rk.pv/sizeBin0+1,base=2))
  rep_helper <- c(table(grp))
  alphaBin = alpha * 2 ^ -(1:nbins) / sizeBin
  alphaBin_dat <- rep(alphaBin, rep_helper)

  # create pvalues and bin variables
  data$step1p <- pchisq(data[, step1_statistic], df = step1_df, lower.tail = F)
  data$step2p <- pchisq(data[, 'lrtgxe'], df = 1, lower.tail = F)
  data                   <- data[order(-data[,step1_statistic]), ]   # sort by step1_statistic
  data$bin_number        <- as.numeric(grp)
  data$step2p_siglvl     <- as.numeric(alphaBin_dat)
  data$log_step2p_siglvl <- -log10(data$step2p_siglvl)
  data$log_step2p        <- -log10(data$step2p)

  significant_hits <- data[which(data$step2p < data$step2p_siglvl), c("snp", "step1p", "step2p")]

  return(significant_hits)
}
```


### Twostep D|G
```{r, echo = T}
plot_twostep(output, step1_statistic = 'lrtdg', step1_df = 1, sizeBin0 = 5, alpha = 0.05)
kable(sig_twostep(output, step1_statistic = 'lrtdg', step1_df = 1, sizeBin0 = 5, alpha = 0.05)) %>% kable_styling()
```

### Twostep E|G
```{r, echo = T}
plot_twostep(output, step1_statistic = 'lrteg', step1_df = 1, sizeBin0 = 5, alpha = 0.05)
kable(sig_twostep(output, step1_statistic = 'lrteg', step1_df = 1, sizeBin0 = 5, alpha = 0.05)) %>% kable_styling()

```

### Twostep EDGE
```{r, echo = T}
plot_twostep(output, step1_statistic = 'lrtedge', step1_df = 2, sizeBin0 = 5, alpha = 0.05)
kable(sig_twostep(output, step1_statistic = 'lrtedge', step1_df = 2, sizeBin0 = 5, alpha = 0.05)) %>% kable_styling()

```


<br>

----

<br>


# Your turn

You will perform GxE analysis to assess interactions between maternal **smoking** + genetic variants in determining childhood BMI using the GxEScanR package. All relevant files are stored in the `bash data` folder.  

Prepare data - specify variables for analysis and obtain variables from the MultiAssay file:
```{r, eval = F, class.source = "fold-show"}
## specify analysis variables
outcome.Name <- "hs_bmi_c_cat"
exposure.Name <- "e3_asmokcigd_p_None"
covariate.Names <- c("h_mbmi_None","e3_sex_None","h_age_None","h_cohort","h_edumc_None","ethn_PC1","ethn_PC2")
snp.Names <- paste("SNP", 1:1000, sep=".")
variables <- c(covariate.Names, exposure.Name, "h_ethnicity_cauc", snp.Names)

## Subset dataset to the variables we want
load("~/Google Drive/My Drive/SHARP.MultiomicsWorkshop.USC.share/data/HELIX.MultiAssayExperiment.RData")
d <- wideFormat(intersectColumns(helix_ma[variables, ,]), colDataCols=outcome.Name) 

# 1) select variables but keep in MultiAssayExperiment format
# 2) intersectionColumns selects only individuals with complete data
# 3) wideFormat returns as a DataFrame

```

<br>

Next, create covariate files, re-arrange column order to suit requirements of the GxEScanR package.    

The order of variables should be as follows: 
* ID
* Outcome
* Covariates 
* Exposure (interaction variable)


```{r, eval = F, class.source = "fold-show"}

```

<br>


Genotype data (VCF + INFO files will be provided rather than generated by the partipants, but just want to emphasize that the file type requirements are the ones commonly output by imputation software.. when analyzing measured genotypes, will likely need to convert plink etc into VCF, which is easy to do)

```{r}



```
