---
title: 'SHARP Multiomics Workshop: Exposome Challenge Data: Pre-Workshop'
author: "David Conti"
date: "`r Sys.time()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: hide
  pdf_document: default
---

```{css, echo=FALSE}
pre {
  max-height: 200px;
  overflow-y: auto;
}
```

```{r setup, include=FALSE, echo=FALSE}
library(knitr)
library(reshape2)
library(ggplot2)
library(epiR)
library(summarytools) # for summarizing variables
library(tidyverse)
library(glmnet)
library(Biobase)
library(gap)

options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE)

# folder for analysis
setwd("~/Google Drive/My Drive/Class/SHARP.MultiomicWorkshop/ExposomeChallengeData/SHARP")

# suppress warning messages for final rendering
old.warn <- getOption("warn")
options(warn=-1)

```

# SHARP MULTIOMICS WORKSHOP
## Exposome Data Challenge
The specific data is from the Exposome Data Analysis Challenge (https://www.isglobal.org/-/exposome-data-analysis-challenge). The Exposome dataset represents a real case scenario of exposome dataset (based on the HELIX project database) with multiple correlated variables (N>100 exposure variables) arising from general and personal environments at different time points, biological molecular data (multi-omics: DNA methylation, gene expression, proteins, metabolomics) and multiple clinical phenotypes. The population is drawn from a multi-center study which will represent the main confounding structure in the dataset.

In addition, for the SHARP Multiomics Workshop, we simulated germline genetics data.

## The HELIX study
The HELIX study represents a collaborative project across six established and ongoing longitudinal population-based birth cohort studies in six European countries (France, Greece, Lithuania, Norway, Spain, and the United Kingdom). HELIX used a multilevel study design with the entire study population totaling 31,472 mother–child pairs, recruited during pregnancy, in the six existing cohorts (first level); a subcohort of 1301 mother-child pairs where biomarkers, omics signatures and child health outcomes were measured at age 6-11 years (second level); and repeat-sampling panel studies with around 150 children and 150 pregnant women aimed at collecting personal exposure data (third level). For more details on the study design see Vrijheid, Slama, et al. EHP 2014. see https://www.projecthelix.eu/index.php/es/data-inventory for more information regarding the study.
<br>


# EXPOSOME ANALYSIS
```{r Exposome: Data Analysis setup, echo=FALSE }

# Outcome
outcome.Name <- "hs_bmi_c_cat" # "hs_asthma" # "hs_bmi_c_cat" "hs_zbmi_who"

# Covariates
# covariate.Names <- c("h_mbmi_None","hs_c_height_None","hs_c_weight_None","hs_wgtgain_None",
#                      "e3_gac_None","e3_sex_None","e3_yearbir_None","h_age_None",
#                      "h_cohort","h_edumc_None","h_native_None","h_parity_None",
#                      "hs_child_age_None")
covariate.Names <- c("h_mbmi_None","e3_sex_None","h_age_None","h_cohort","h_edumc_None")

# Exposure related
exposure.group <- "Organochlorines" # {"Metals", "Organochlorines", "Organophosphate pesticides", "PBDE", "PFAS", "Phenols", "Phthalates", "All"}

# All potential exposures ategorized by groups
S <- 7
exposure.Names <- list(rep(0,S))
exposure.Names[[1]] <- c("hs_as_m_Log2","hs_cd_m_Log2","hs_co_m_Log2","hs_cs_m_Log2",
                          "hs_cu_m_Log2","hs_hg_m_Log2","hs_mn_m_Log2","hs_mo_m_Log2",
                          "hs_pb_m_Log2")
exposure.Names[[2]] <- c("hs_dde_madj_Log2","hs_ddt_madj_Log2","hs_hcb_madj_Log2","hs_pcb118_madj_Log2",
                          "hs_pcb138_madj_Log2","hs_pcb153_madj_Log2","hs_pcb170_madj_Log2","hs_pcb180_madj_Log2")
exposure.Names[[3]] <- c("hs_dep_madj_Log2","hs_detp_madj_Log2","hs_dmp_madj_Log2","hs_dmtp_madj_Log2")
exposure.Names[[4]] <- c("hs_pbde153_madj_Log2","hs_pbde47_madj_Log2")
exposure.Names[[5]] <- c("hs_pfhxs_m_Log2","hs_pfna_m_Log2","hs_pfoa_m_Log2","hs_pfos_m_Log2","hs_pfunda_m_Log2")
exposure.Names[[6]] <- c("hs_bpa_madj_Log2","hs_bupa_madj_Log2","hs_etpa_madj_Log2","hs_mepa_madj_Log2",
                          "hs_oxbe_madj_Log2","hs_prpa_madj_Log2","hs_trcs_madj_Log2")
exposure.Names[[7]] <- c("hs_mbzp_madj_Log2","hs_mecpp_madj_Log2","hs_mehhp_madj_Log2","hs_mehp_madj_Log2",
                          "hs_meohp_madj_Log2","hs_mep_madj_Log2","hs_mibp_madj_Log2","hs_mnbp_madj_Log2",
                          "hs_ohminp_madj_Log2","hs_oxominp_madj_Log2")

# Analysis models to run
univariate <- T
ridge <- T
lasso <- T
elasticnet <- T

```

## Exposome: Overview
Analysis of exposure/chemical mixtures within a single group of related exposures

### The Question of interest:
- How are mixtures of exposures for **`r exposure.group`** associated with the outcome **`r outcome.Name`**?
  <br>

### Exposure mixture analysis
Often in assessing multiple exposures we have several questions or goals interest:  
1) what is the independent effect of each exposure?
2) do combinations of exposures act in a synergistic manner to increase risk? and,
3) what is the combined effect when an individual is exposed to a mixture of compounds?

### Generalized linear regression: Defining the outcome and the linear predictor
Using generalized linear regression models, we assume that the outcome, $Y$, is generated from a specific distribution of the exponential family and that the mean $\mu$ of this distribution is a function $\eta$ of the $P$ exposures defined in the design matrix $\boldsymbol{X}$. Specifically, $E[Y|\boldsymbol{X}] = \mu = g^{-1}(f(\boldsymbol{X}))$, where $g(.)$ is a link function connecting the function of the exposures $\eta$  Common link functions include the identify link for the normal distribution resulting in linear regression and a logit link for a binomial distribution resulting in logistic regression. Defined on the scale of the outcome as transformed by the link function, combinations of exposures act additivity. 

$\eta = \boldsymbol{X\beta}$) to the mean of the distribution, $g(\mu) = \boldsymbol{X\beta}=\eta$.


## Exposome: Processing the Data
```{r Exposome: Processing the Data, echo=TRUE}
load("../exposome.RData")

# Merge all data by ID
d <- merge(x=phenotype, covariates, by.x="ID", by.y="ID", all=T)
d <- merge(x=d, y=exposome, by.x="ID", by.y="ID", all=T)

# Create exposure design matrix
if(exposure.group!="All") {
  exposure.index <- switch(exposure.group, 
            "Metals"=1,
            "Organochlorines"=2,
            "Organophosphate pesticides"=3,
            "PBDE"=4,
            "PFAS"=5,
            "Phenols"=6,
            "Phthalates"=7)
  exposure.Names <- unlist(exposure.Names[[exposure.index]]) # focus on single group
}
if(exposure.group=="All") { exposure.Names <- unlist(exposure.Names) } # treat all groups as a single group
# create exposure matrix X and XtX matrix for g-prior
X <- scale(as.matrix(d[,exposure.Names]), center=T, scale=T)


# Create the outcome variable
Y <- d[,outcome.Name] # outcome
if(outcome.Name=="hs_bmi_c_cat") { Y <- ifelse(as.numeric(Y)>=3, 1, 0)}

# Create the covariate design matrix
U <- as.data.frame(d[,covariate.Names])
U[,"h_cohort"] <- as.factor(U[,"h_cohort"])
U[,"h_edumc_None"] <- as.factor(U[,"h_edumc_None"])
U <- model.matrix(as.formula(paste("~-1+", paste(c(covariate.Names), collapse="+"))), data=U)
#U <- scale(U, center=T, scale=F) # mean center

# Other variables for analysis
N <- nrow(d) # number of individuals in the analysis
Q <- ncol(U)  # number of covariates in the matrix U
P <- ncol(X)  # number of expsoures in the matrix X

```
<br>

## Exposome: Descriptive Statistics for `r exposure.group`: {.tabset}
- Exposures include a total of `r P` exposures and include the following exposures:  
`r exposure.Names` 

### Summary Table for Each Exposure
```{r Exposome: summmary statistics, echo=TRUE}
summarytools::view(dfSummary(as.data.frame(X), style = 'grid',
                               max.distinct.values = 10, plain.ascii =   FALSE, valid.col = FALSE, headings = FALSE), method = "render")

```

### Correlation Matrix for the Exposures:
```{r Exposome: cor.plot, echo=TRUE}
cormat <- round(cor(X, use="complete.obs"), 2)
cormat[lower.tri(cormat)]<- NA
melted_cormat <- melt(cormat)
ggplot(data = melted_cormat, aes(Var2, Var1, fill = value))+
  geom_tile(color = "white")+
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1), space = "Lab", 
                       name="Pearson\nCorrelation") +
  theme_minimal()+ 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                   size = 12, hjust = 1))+
  coord_fixed() + geom_text(aes(Var2, Var1, label = value), color = "black", size = 2)

```
<br>

## Exposome: Statistical Approaches for Association Analysis
### Univariate Regression
#### Univariate Model:
We first investigate the independent association of each of $P$ exposures on the outcome. thus for $p = {1,...,P}$ we fit the following model:

$y_{i} \sim Bernoulli(\mu_{ip})$
  
$logit(\mu_{ip}) = \alpha_{p} + \beta_{p}X_{p} + \sum_Q{\delta_{qp}U_q}$
  
Each effect estimate $\beta_{p}$ is estimated by a maximum likelihood estimats (MLE), $\hat{\beta}$. Note that the $X_{p}$ are mean centered.

```{r Exposome: Univariate model, echo=TRUE}
if(univariate) {
  univariate.results <- t(sapply(1:P, FUN=function(p) {  # using index p facilitate write
    x <- X[,p]
    reg <- glm(Y~x+U, family=binomial)    # perform logistic regression
    s.reg <- summary(reg)                 # get the summary for the regression
    c.reg <- s.reg$coef["x",]             # select the coefficients for the exposure
    write.table(t(c(exposure.Names[p], c.reg)), file="ExposomeUnivariateResults.txt", append=ifelse(p==1, F, T), quote=F, sep="\t", col.names=ifelse(p==1, T, F), row.names=F)
    return(c.reg)                         # to avoid potential memory issues only return coefficients if small number of exposures
  }, simplify=T))
  univariate.results <- data.frame(exposure.Names,univariate.results)
}

```

#### Univariate results: {.tabset}
##### Univariate Summary Table:
```{r Exposome: Univariate table}
if(univariate) { kable(univariate.results, digits=3, align="c", row.names=FALSE, col.names=c("Exposure","Estimate", "SD","Lower 95% CI", "Upper 95% CI"))}
```

##### Univariate Manhattan Plot:
```{r Exposome: Univariate plot}
neglog.pvalues <- -log10(univariate.results$Pr...z..)
plot(1:nrow(univariate.results), neglog.pvalues, 
     pch=16, xaxt="n", ylim=c(0, max(neglog.pvalues, 3)),
     ylab="-log(p-value", xlab="")
text(x=1:nrow(univariate.results), y=par("usr")[3]-0.1, xpd=NA,
     labels=univariate.results$exposure.Names, adj=.9, srt=45, cex=.75)
abline(h=-log10(0.05/nrow(univariate.results)), lty=2, lwd=2, col=2)
```
<br>

### Ridge regression:
#### Ridge Model:
To estimate the independent effect of each exposure, we  model all exposures jointly in a single regression model:

$y_{i} \sim Bernoulli(\mu_{i})$
  
$logit(\mu_{i}) = \alpha + \sum_P{\beta_{p}X_{p}} + \sum_Q{\delta_{qp}U_q}$

The $\beta$s are specified with a single normal prior with a joint variance: $\beta_{p} \sim N(0, \sigma^2)$. This approach is referred to as ridge regression. Posterior estimates $\tilde{\beta}$ are a weighted average between the MLE $\hat{\beta}$ obtained and the prior mean, $\bar{\beta}=0$:

$\tilde{\beta}=W\bar{\beta} + (1-W)\hat{\beta}$ where the weight is a balance between the estimated uncertainty for the MLE $V_{\hat{\beta}}$, which reflects the information contained by the data, and the estimated common prior variance, $\sigma^{2}$: $W=V_{\hat{\beta}}/(\sigma^{2}+V_{\hat{\beta}})$. Thus, if the uncertainty in the MLE is large relative to estimated group variance, then the posterior estimate is shrunk towards zero. As precision increases, the posterior estimate will be weighted more towards the MLE estimates.

NEED TO ADD OPTIMIZATION interpretation...

Note that replacing the prior on the $\beta$s with a double exponential induces sparsity and is Bayesian implementation of the lasso model. A mixture prior between a normal and double exponential is similar to elastic net. 

```{r Exposome: Ridge Regression, echo=TRUE}
if(ridge) {
  ridge.cv <- cv.glmnet(x=X, y=Y, family="binomial", alpha=0)  # alpha=0 is for ridge
  ridge.coef <- coef(ridge.cv, s = "lambda.min")
  ridge.fit <- glmnet(x=X, y=Y, family="binomial", alpha=0)
}
```

#### Ridge Results: {.tabset}
##### Ridge Selection of $\lambda$ via Cross Validation
```{r Exposome: Ridge Cross Validation}
if(ridge) { plot(ridge.cv) }
```

##### Ridge Coefficient Shrinkage
```{r Exposome: Ridge Shrinkage}
if(ridge) { 
  plot(ridge.fit, xvar="lambda", label=T)
  abline(v=log(ridge.cv$lambda.min), lty=2, col="red")
  abline(v=log(ridge.cv$lambda.1se), lty=2, col="green")
}
```

##### Ridge Coefficients for the Selected Model
```{r Exposome: Ridge coefficients}
if(ridge) { ridge.coef }
```

<br>


### LASSO regression:
#### LASSO Model:

Note that replacing the prior on the $\beta$s with a double exponential induces sparsity and is Bayesian implementation of the lasso model. A mixture prior between a normal and double exponential is similar to elastic net. 

```{r Exposome: LASSO Regression, echo=TRUE}
if(lasso) {
  lasso.cv <- cv.glmnet(x=X, y=Y, family="binomial", alpha=1)  # alpha=1 is for lasso
  lasso.coef <- coef(lasso.cv, s = "lambda.min")
  lasso.fit <- glmnet(x=X, y=Y, family="binomial", alpha=1)
}
```

#### LASSO Results: {.tabset}
##### LASSO Selection of $\lambda$ via Cross Validation
```{r Exposome: LASSO Cross Validation}
if(lasso) { plot(lasso.cv) }
```

##### LASSO Coefficient Shrinkage
```{r Exposome: LASSO Shrinkage}
if(lasso) { 
  plot(lasso.fit, xvar="lambda", label=T)
  abline(v=log(lasso.cv$lambda.min), lty=2, col="red")
  abline(v=log(lasso.cv$lambda.1se), lty=2, col="green")
}
```

##### LASSO Coefficients for the Selected Model
```{r Exposome: LASSO coefficients}
if(lasso) { lasso.coef }
```
<br>

### Elastic net regression:
#### Elastic net Model:

A mixture prior between a normal and double exponential is similar to elastic net. 

```{r Exposome: Elastic net Regression, echo=TRUE}
if(elasticnet) {
  elasticnet.cv <- cv.glmnet(x=X, y=Y, family="binomial", alpha=0.5)  # alpha=0.5 is for elastic net
  elasticnet.coef <- coef(elasticnet.cv, s = "lambda.min")
  elasticnet.fit <- glmnet(x=X, y=Y, family="binomial", alpha=0.5)
}
```

#### Elastic net Results: {.tabset}
##### Elastic net Selection of $\lambda$ via Cross Validation
```{r Exposome: Elastic net Cross Validation}
if(elasticnet) { plot(elasticnet.cv) }
```

##### Elastic net Coefficient Shrinkage
```{r Exposome: Elastic net Shrinkage}
if(elasticnet) { 
  plot(elasticnet.fit, xvar="lambda", label=T)
  abline(v=log(elasticnet.cv$lambda.min), lty=2, col="red")
  abline(v=log(elasticnet.cv$lambda.1se), lty=2, col="green")
}
```

##### Elastic net Coefficients for the Selected Model
```{r Exposome: Elastic net coefficients}
if(elasticnet) { elasticnet.coef }
```
<br>

***

# PROTEOME ANALYSIS
```{r Proteome: Data Analysis setup, echo=FALSE }

# Outcome
outcome.Name <- "hs_bmi_c_cat" # "hs_asthma" # "hs_bmi_c_cat" "hs_zbmi_who"

# Covariates
covariate.Names <- c("e3_sex_None","age_sample_years","ethn_PC1","ethn_PC2","hs_dift_mealblood_imp","blood_sam4")


# Analysis models to run
univariate <- T
ridge <- T
lasso <- T
elasticnet <- T

```

## Proteome: Processing the Data
```{r Proteome: Processing the Data, echo=TRUE}
load("../exposome.RData")
load("../proteome.RData")

# Merge all data by ID
d <- merge(x=phenotype, covariates, by.x="ID", by.y="ID", all=T)
d <- merge(x=d, y=exposome, by.x="ID", by.y="ID", all=T)
d.proteome <- data.frame(proteome@phenoData@data, t((proteome@assayData$exprs))) # assumes data are in exactly same order -->
d <- merge(x=d, y=d.proteome, by.x="ID", by.y="ID", all=T)
d <- d[complete.cases(d), ] # helps to facilitate analysis

# Create exposure design matrix
proteome.Names <- fData(proteome)$Prot_ID
X <- d[,proteome.Names]
# create exposure matrix X
X <- scale(X, center=T, scale=T)

# Create the outcome variable
Y <- d[,outcome.Name] # outcome
if(outcome.Name=="hs_bmi_c_cat") { Y <- ifelse(as.numeric(Y)>=3, 1, 0)}

# Create the covariate design matrix
U <- as.data.frame(d[,covariate.Names])
U <- model.matrix(as.formula(paste("~-1+", paste(c(covariate.Names), collapse="+"))), data=U)
#U <- scale(U, center=T, scale=F) # mean center

# Other variables for analysis
N <- nrow(d) # number of individuals in the analysis
Q <- ncol(U)  # number of covariates in the matrix U
P <- ncol(X)  # number of expsoures in the matrix X

```
<br>

## Proteome: Descriptive Statistics {.tabset}  
- The proteome includes a total of `r P` proteomic measures and include the following:  
`r proteome.Names`

### Summary Table for Each Proteomic Feature
```{r Proteome: summmary statistics, echo=TRUE}
summarytools::view(dfSummary(as.data.frame(X), style = 'grid',
                               max.distinct.values = 10, plain.ascii =   FALSE, valid.col = FALSE, headings = FALSE), method = "render")

```

### Correlation Matrix for the Proteome:
```{r Proteome: cor.plot, echo=TRUE}
cormat <- round(cor(X, use="complete.obs"), 2)
cormat[lower.tri(cormat)]<- NA
melted_cormat <- melt(cormat)
ggplot(data = melted_cormat, aes(Var2, Var1, fill = value))+
  geom_tile(color = "white")+
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1), space = "Lab", 
                       name="Pearson\nCorrelation") +
  theme_minimal()+ 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                   size = 6, hjust = 1))+
  coord_fixed()
  #+geom_text(aes(Var2, Var1, label = value), color = "black", size = 2)

```
<br>

## Proteome: Statistical Approaches for Association Analysis
### Univariate Regression

```{r Proteome: Univariate model, echo=TRUE}
if(univariate) {
  univariate.results <- t(sapply(1:P, FUN=function(p) {  # using index p facilitate write
    x <- X[,p]
    reg <- glm(Y~x+U, family=binomial)    # perform logistic regression
    s.reg <- summary(reg)                 # get the summary for the regression
    c.reg <- s.reg$coef["x",]             # select the coefficients for the exposure
    write.table(t(c(proteome.Names[p], c.reg)), file="ProteomeUnivariateResults.txt", append=ifelse(p==1, F, T), quote=F, sep="\t", col.names=ifelse(p==1, T, F), row.names=F)
    return(c.reg)                         # to avoid potential memory issues only return coefficients if small number of exposures
  }, simplify=T))
  univariate.results <- data.frame(proteome.Names,univariate.results)
}

```

#### Univariate results: {.tabset}
##### Univariate Summary Table:
```{r Proteome: Univariate table}
if(univariate) { kable(univariate.results, digits=3, align="c", row.names=FALSE, col.names=c("Protein","Estimate", "SD","Lower 95% CI", "Upper 95% CI"))}
```

##### Univariate Manhattan Plot:
```{r Proteome: Univariate plot}
neglog.pvalues <- -log10(univariate.results$Pr...z..)
plot(1:nrow(univariate.results), neglog.pvalues, 
     pch=16, xaxt="n", ylim=c(0, max(neglog.pvalues, 3)),
     ylab="-log(p-value", xlab="")
text(x=1:nrow(univariate.results), y=par("usr")[3]-0.1, xpd=NA,
     labels=univariate.results$proteome.Names, adj=.9, srt=45, cex=.75)
abline(h=-log10(0.05/nrow(univariate.results)), lty=2, lwd=2, col=2)
```
<br>

### Ridge regression:
#### Ridge Model:

```{r Proteome: Ridge Regression, echo=TRUE}
if(ridge) {
  ridge.cv <- cv.glmnet(x=X, y=Y, family="binomial", alpha=0)  # alpha=0 is for ridge
  ridge.coef <- coef(ridge.cv, s = "lambda.min")
  ridge.fit <- glmnet(x=X, y=Y, family="binomial", alpha=0)
}
```

#### Ridge Results: {.tabset}
##### Ridge Selection of $\lambda$ via Cross Validation
```{r Proteome: Ridge Cross Validation}
if(ridge) { plot(ridge.cv) }
```

##### Ridge Coefficient Shrinkage
```{r Proteome: Ridge Shrinkage}
if(ridge) { 
  plot(ridge.fit, xvar="lambda", label=T)
  abline(v=log(ridge.cv$lambda.min), lty=2, col="red")
  abline(v=log(ridge.cv$lambda.1se), lty=2, col="green")
}
```

##### Ridge Coefficients for the Selected Model
```{r Proteome: Ridge coefficients}
if(ridge) { ridge.coef }
```

<br>

### LASSO regression:
#### LASSO Model:

```{r Proteome: LASSO Regression, echo=TRUE}
if(lasso) {
  lasso.cv <- cv.glmnet(x=X, y=Y, family="binomial", alpha=1)  # alpha=1 is for lasso
  lasso.coef <- coef(lasso.cv, s = "lambda.min")
  lasso.fit <- glmnet(x=X, y=Y, family="binomial", alpha=1)
}
```

#### LASSO Results: {.tabset}
##### LASSO Selection of $\lambda$ via Cross Validation
```{r Proteome: LASSO Cross Validation}
if(lasso) { plot(lasso.cv) }
```

##### LASSO Coefficient Shrinkage
```{r Proteome: LASSO Shrinkage}
if(lasso) { 
  plot(lasso.fit, xvar="lambda", label=T)
  abline(v=log(lasso.cv$lambda.min), lty=2, col="red")
  abline(v=log(lasso.cv$lambda.1se), lty=2, col="green")
}
```

##### LASSO Coefficients for the Selected Model
```{r Proteome: LASSO coefficients}
if(lasso) { lasso.coef }
```
<br>

### Elastic net regression:
#### Elastic net Model:

```{r Proteome: Elastic net Regression, echo=TRUE}
if(elasticnet) {
  elasticnet.cv <- cv.glmnet(x=X, y=Y, family="binomial", alpha=0.5)  # alpha=0.5 is for elastic net
  elasticnet.coef <- coef(elasticnet.cv, s = "lambda.min")
  elasticnet.fit <- glmnet(x=X, y=Y, family="binomial", alpha=0.5)
}
```

#### Elastic net Results: {.tabset}
##### Elastic net Selection of $\lambda$ via Cross Validation
```{r Proteome: Elastic net Cross Validation}
if(elasticnet) { plot(elasticnet.cv) }
```

##### Elastic net Coefficient Shrinkage
```{r Proteome: Elastic net Shrinkage}
if(elasticnet) { 
  plot(elasticnet.fit, xvar="lambda", label=T)
  abline(v=log(elasticnet.cv$lambda.min), lty=2, col="red")
  abline(v=log(elasticnet.cv$lambda.1se), lty=2, col="green")
}
```

##### Elastic net Coefficients for the Selected Model
```{r Proteome: Elastic net coefficients}
if(elasticnet) { elasticnet.coef }
```
<br>

***

# GENOME ANALYSIS
```{r Genome: Data Analysis setup, echo=FALSE }

# Outcome
outcome.Name <- "hs_bmi_c_cat" # "hs_asthma" # "hs_bmi_c_cat" "hs_zbmi_who"

# Covariates
covariate.Names <- c("h_mbmi_None","e3_sex_None","h_age_None","h_cohort","h_edumc_None","ethn_PC1","ethn_PC2") 

# Analysis models to run
univariate <- T


```

## Genome: Processing the Data
```{r Genome: Processing the Data, echo=TRUE}
load("../exposome.RData")
load("../proteome.RData")
load("../genome.RData")

# Merge all data by ID
d <- merge(x=phenotype, covariates, by.x="ID", by.y="ID", all=T)
d <- merge(x=d, y=exposome, by.x="ID", by.y="ID", all=T)
d <- data.frame(d, G)  # assumes that order of individuals in d and G are the same
d.proteome <- data.frame(proteome@phenoData@data, t((proteome@assayData$exprs))) # assumes data are in exactly same order -->
d <- merge(x=d, y=d.proteome, by.x="ID", by.y="ID", all=T)

d <- d[complete.cases(d), ] # helps to facilitate analysis

# Create  design matrix
snp.Names <- paste("SNP", 1:ncol(G), sep=".")
X <- d[,snp.Names]

# Create the outcome variable
Y <- d[,outcome.Name] # outcome
if(outcome.Name=="hs_bmi_c_cat") { Y <- ifelse(as.numeric(Y)>=3, 1, 0)}

# Create the covariate design matrix
U <- as.data.frame(d[,covariate.Names])
U <- model.matrix(as.formula(paste("~-1+", paste(c(covariate.Names), collapse="+"))), data=U)
#U <- scale(U, center=T, scale=F) # mean center

# Other variables for analysis
N <- nrow(d) # number of individuals in the analysis
Q <- ncol(U)  # number of covariates in the matrix U
P <- ncol(X)  # number of exposures in the matrix X

```
<br>

## Genome: Descriptive Statistics {.tabset}  
- The genome includes a total of `r P` single nucleotide polymorphisms (SNPs):  

### Plot of Genetic Ancestry as Estimated by Prinicpal Components
```{r Genome: PC plot, echo=TRUE}
plot(d$ethn_PC1, d$ethn_PC2, pch=16, col=ifelse(d$h_ethnicity_cauc=="yes", 1, 2),
     xlab="Component 1", ylab="Component 2")
legend(x="topleft", legend=c("Caucasian", "Other"), col=c(1,2), pch=16)

```

### Correlation Matrix for Local Region of the Genome:
```{r Genome: cor.plot, echo=TRUE}
cormat <- round(cor(X[,1:(P/5)], use="complete.obs"), 2)
cormat[lower.tri(cormat)]<- NA
melted_cormat <- melt(cormat)
ggplot(data = melted_cormat, aes(Var2, Var1, fill = value))+
  geom_tile(color = "white")+
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1), space = "Lab", 
                       name="Pearson\nCorrelation") +
  theme_minimal()+
  theme(axis.text.x = element_blank(), axis.text.y = element_blank())+
  labs(y= "SNPs", x = "SNPs")+
  coord_fixed()

```
<br>

## Genome: Statistical Approaches for Association Analysis
### Univariate Regression

```{r Genome: Univariate model, echo=TRUE}
if(univariate) {
  univariate.results <- t(sapply(1:P, FUN=function(p) {  # using index p facilitate write
    x <- X[,p]
    reg <- glm(Y~x+U, family=binomial)    # perform logistic regression
    s.reg <- summary(reg)                 # get the summary for the regression
    c.reg <- s.reg$coef["x",]             # select the coefficients for the exposure
    write.table(t(c(snp.Names[p], c.reg)), file="GenomeUnivariateResults.txt", append=ifelse(p==1, F, T), quote=F, sep="\t", col.names=ifelse(p==1, T, F), row.names=F)
    return(c.reg)                         # to avoid potential memory issues only return coefficients if small number of exposures
  }, simplify=T))
  univariate.results <- data.frame(snp.Names,univariate.results)
}

```

#### Univariate results: {.tabset}
##### Univariate Summary Table:
```{r Genome: Univariate table}
if(univariate) { kable(univariate.results[univariate.results$Pr...z..<0.05,], digits=3, align="c", row.names=FALSE, col.names=c("Protein","Estimate", "SD","Lower 95% CI", "Upper 95% CI"))}
```

##### Univariate Manhattan Plot:
```{r Genome: Univariate MH plot}
neglog.pvalues <- -log10(univariate.results$Pr...z..)
plot(1:nrow(univariate.results), neglog.pvalues, 
     pch=16, xaxt="n", ylim=c(0, max(neglog.pvalues, 3)),
     ylab="-log(p-value)", xlab="SNPs")
abline(h=-log10(0.05/nrow(univariate.results)), lty=2, lwd=2, col=2)
```

##### Univariate QQ-Plot:
```{r Genome: QQ-plot}
pvalues <- univariate.results$Pr...z..
r <- gcontrol2(pvalues, pch=16)
lambda <- round(r$lambda,3)
text(x=1, y=5, labels=bquote(lambda == .(lambda)), cex=2)

```

<br>

```{r final clean up}
options(warn=old.warn)

```